Hydraulically-vulnerable trees survive on deep-water access in a tropical forest community

### 2020-10-29 manuscript.rmd
1. Corrected erd equation 2. Other minor changes

\
  Trade-offs in plant 'traits' shape how plants make a living in their environment leading to alternative ecological strategies. Characterizing the essential constraints necessary to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019] entails identifying the topography of trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality). Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates. These efforts have resulted in identification of correlations between above-ground hydraulic traits such as vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016]. Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plants' realized water environments, specifically plant's ability to acquire water via differing rooting depths in hydrological rather than meteorological droughts, are rarely studied, especially in the tropical region. We tested how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tree species of a tropical forest over 35 years that witnessed droughts of various intensity and duration.

@@

We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower ($\cong$ $E_{v} + E_{g}$; 2012-2017; see [SI](#SIData.matteo))), (ii) stream discharge ($q_\textrm{run}$) from the Conrad catchment ($\cong$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$) $\cong$ $W_{\textrm{liq,} z}$ from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical time domain reflectometry (TDR) probe over 0-15 cm depth ($\mathrm{\theta}_\textrm{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter,  $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).

@@

\begin{equation}
\textrm{ERD}_{s} = \underset{\forall \, h \, \in \, H}{\textrm{median}} \Big\{z,h\,\Big|\, \Big(\textrm{L}_{s|z,h} = \underset{\forall \, z \,\in\,Z}{\textrm{maximum}}(\textrm{L}_{s|~z,h})\Big) \cap \Big(r_{s|z,h} \ge 0.71\Big)\Big\}
(\#eq:erd)
\end{equation}
where $L$ is the likelihood, $H$ = {1, .., 100}, refer to the 100 hydrological realizations and $Z$ = {0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m. We removed finely resolved soil layers below 0.21 m and averaged $\mathrm{\Psi}_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 at 0.4 m as sufficient resolution for Model validation.

@@

## Conclusions
\
Establishing relationships between environment, traits and demographic outcomes of plants is imperative for developing a predictive plant ecology. Trees realized water environments and water-access are nonetheless rarely studied. To our knowledge, this is the first study to demonstrate a mechanistic link between realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a tree community. We report here that deep-water access plays a role in alleviating mortality of hydraulically vulnerable trees. This has important implications for our predictive understanding of tropical forest dynamics under current and future climate. Our community-scale framework for modeling effective rooting depths and leaf vulnerability curves indicates the possibilities in expanding the use of these critical, hard-earned observations in species-rich forests towards community-scale generalizations.

@@

\
The $K_h$ segments were removed from the branches by repeatedly cutting back the branch on both ends while it was submerged in water to relax the hydraulic tension within the xylem ((REFERENCE)). Then the ends were shaved clean with a fresh razor. Segments were 11–34 cm in length (mean ± SD = 22 ± 4 cm) and 0.2–1.6 cm in diameter (0.7 ± 0.2 cm). The segments were placed in a hydraulic conductivity apparatus similar to that described by Sperry et al. (1998) except that flow rate was measured with graduated pipettes instead of a balance. The perfusion solution was 10 mM KCl filtered to 0.2 mM. For each segment, we measured the flow rate at three pressure heads: two that ranged 0.5–2 kPa and at zero pressure. Following Torres-Ruiz et al. (2012), $K_h$ was calculated as the slope of the flow rate versus pressure gradient (i.e., pressure head divided by segment length) among the three pressure heads. Stem area specific hydraulic conductivity was ($K_\textrm{stem}$) was calculated by dividing $K_h$ by the cross-sectional area of stem. To calculate cross-sectional stem area, stem diameter was measured twice perpendicularly on both ends of the segment and averaged, then area was calculated assuming a circular cross section.

@@

```{r sens-sm-et, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-sm-et}Parameter sensitivity of ELM-FATES average monthly soil water content (that is, VWC, a-b) and ET (c-d) is shown for two representative sample months, as sensitivity for a particular parameter varies depending on the chosen month. Horizontal axes show global ranges for a given parameter. Parameters are color coded and are consistent across panels. Each point represents a simulation among the 5000 ensemble-member runs. See S\\@ref(tab:param) for description, global ranges and chosen ranges for parameters."}
knitr::include_graphics(file.path("figures/ELM_FATES_parameter_sensitivity/sens.sm.et.jpeg"))
```

>>>>

\
  Trade-offs in plant 'traits' shape how plants make a living in their environment leading to alternative ecological strategies. Characterizing the essential constraints necessary to model species-rich communities under a changing environment [@Koven:2020; @Christoffersen:2016kd2d; @Wright:2010; @Bartlett.2019] entails identifying the topography of trait trade-offs in different environments, and how these relate to demographic rates (growth, recruitment and mortality). Significant efforts have been invested into identifying and linking universal drought indices, above-ground traits, and demographic rates. These efforts have resulted in identification of correlations between above-ground hydraulic traits such as vulnerability to embolism and hydraulic safety margins and mortality [@Anderegg:2016], although numerous counter examples also exist [@Hoffmann.2011; @Johnson:2018fad; @Nardini:2015; @Paddock:2013, @Venturas.2016]. Deep roots may mitigate hydraulic vulnerability and mortality-risk, yet plant's ability to acquire water via differing rooting depths and the actual water-availability by depth during droughts are rarely studied, especially in the tropical region. We tested how rooting depth is linked to above-ground hydraulic traits and mortality for co-occurring tree species of a tropical forest over 35 years that witnessed droughts of various intensity and duration.

@@

We calibrated ELM-FATES over 2012-2018 against three key fluxes and states in Eq. \@ref(eq:water-balance), namely, (i) evapotranspiration (ET) from the eddy flux tower ($\cong$ $E_{v} + E_{g}$; 2012-2017; see [SI](#SIData.matteo))), (ii) stream discharge ($q_\textrm{run}$) from the Conrad catchment ($\cong$ $q_\textrm{over} + q_\textrm{drai}$; 2012-2018; see [SI](#SIData.steveP)) and (iii) soil volumetric water content (VWC or $\theta_{z}$ $\cong$ $W_{\textrm{liq,} z}$) from two sources: a long-term (2012-2018) record of VWC from a point location near the eddy flux tower we obtained with a vertical time domain reflectometry (TDR) probe over 0-15 cm depth ($\mathrm{\theta}_\textrm{0-0.15 m}$) and existing 50 ha plot-wide snap-shot measurements of VWC during the dry season of 2015 and 2016 for the depths of 0.15, 0.4 and 1 m with 1299 samples in total covering all soil types and habitats (see [SI](#SIData.VWC) and @Kupers:2019 for details; hereafter,  $\mathrm{\theta}_{\textrm{0-0.15 m}}$, $\mathrm{\theta}_{\textrm{0.4 m}}$ and $\mathrm{\theta}_{\textrm{1 m}}$ refer to plot-wide averages).

@@

\begin{equation}
\textrm{ERD}_{s} = \underset{\forall \, h \, \in \, H}{\textrm{median}} \Big(z\,\Big|\, \textrm{L}_{s|z,h} = \underset{\forall \, z \,\in\,Z}{\textrm{maximum}}(\textrm{L}_{s|z,h}| r_{s|z,h} \ge 0.71\Big)\Big)
(\#eq:erd)
\end{equation}
where $L$ is the likelihood of $z$ given the observed growth time series, $H$ = {1, .., 100}, refer to the 100 hydrological realizations and $Z$ = {0.4, `r paste(soil.depths[8:length(soil.depths)], collapse=", ")`} m. We removed finely resolved soil layers below 0.21 m and averaged $\mathrm{\Psi}_{\textrm{soil},z}$ for depths 0.21, 0.37 and 0.62 at 0.4 m as sufficient resolution for Model validation.

@@

## Conclusions
\
Establishing relationships between environment, traits and demographic outcomes of plants is imperative for developing a predictive plant ecology. Trees realized water environments and water-access are nonetheless rarely studied. To our knowledge, this is the first study to demonstrate a mechanistic link between realized drought experience, above-ground hydraulic architecture and long-term mortality outcomes for a tree community. We report here that deep-water access plays a role in alleviating mortality of hydraulically vulnerable trees. This has important implications for our predictive understanding of tropical forest dynamics under current and future climate. Our community-scale framework for modeling effective rooting depths and leaf vulnerability curves indicates the possibilities in expanding the use of these critical, rare observations in species-rich forests towards community-scale generalizations.

@@

\
The $K_h$ segments were removed from the branches by repeatedly cutting back the branch on both ends while it was submerged in water to relax the hydraulic tension within the xylem ((REFERENCE)). Then the ends were shaved clean with a fresh razor. Segments were 11–34 cm in length (mean ± SD = 22 ± 4 cm) and 0.2–1.6 cm in diameter (0.7 ± 0.2 cm). The segments were placed in a hydraulic conductivity apparatus similar to that described by Sperry et al. (1998) except that flow rate was measured with graduated pipettes instead of a balance. The perfusion solution was 10 mM KCl filtered to 0.2 mM. For each segment, we measured the flow rate at three pressure heads: two that ranged 0.5–2 kPa and at zero pressure. Following Torres-Ruiz et al. (2012), $K_h$ was calculated as the slope of the flow rate versus pressure gradient (i.e., pressure head divided by segment length) among the three pressure heads. Stem area specific hydraulic conductivity was ($K_\textrm{stem}$) was calculated by dividing $K_h$ by the cross-sectional area of the stem. To calculate cross-sectional stem area, stem diameter was measured twice perpendicularly on both ends of the segment and averaged, then area was calculated assuming a circular cross section.

@@

```{r sens-sm-et, eval = TRUE, echo = FALSE, message = FALSE, fig.align ='center', out.width = "70%", fig.pos = 'H', results = "asis", fig.cap = "\\label{lab-sens-sm-et}Parameter sensitivity of ELM-FATES soil water content (that is, VWC, a-b) and ET (c-d) is shown for two representative sample months, as sensitivity for a particular parameter varies depending on the chosen month. Horizontal axes show global ranges for a given parameter. Parameters are color coded and are consistent across panels. Each point represents a simulation among the 5000 ensemble-member runs. See S\\@ref(tab:param) for description, global ranges and chosen ranges for parameters."}
knitr::include_graphics(file.path("figures/ELM_FATES_parameter_sensitivity/sens.sm.et.jpeg"))
```
